Topological Equivariant Artist Model

### 2021-03-31 paper/paper.tex
paper paper

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
%% this is still a mess, come back to it
Models for describing visualizations and visualization tools are deeply tied to to the domains in which the visualizations are made \cite{friendlyBriefHistoryData2006a,pousmanCasualInformation2007}, and these tools are often made from building block libraries \cite{wongsuphasawatNavigatingWideWorld2021} such as the Python visualization library Matplotlib. Because building block libraries often support so many domains, the API can grow organically in ways that are incoherant and brittle. To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization, however it is specified in the library built out of these blocks, is guaranteed to be a faithful representation of the data. Our model, which we call the  topological equivariant artist model (TEAM), provides a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.

\begin{figure}
  \includegraphics[width=\columnwidth]{figures/continuity.png}
  \caption{Continuity describes how the records in the dataset, in this case the rows in the table, are connected to each other. The line plot is an invalid visual representation of the table because it implies that the records are connected to each other in some way. On the other hand the bar plot is valid because it preserves the records as independent data points.}
  \label{fig:intro:continuity}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figures/equivariance.png}
  \caption{Equivariance means that transformations on the data side and on the visual side are equal. In this example, the temperature is doubled via the arrows representing the doubling action, and the plot of the temperature is doubled via the arrows on the visual side.}
  \label{fig:intro:equivariance}
\end{figure}

Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation. The contribution of TEAM is 
\begin{enumerate}
  \item topology preserving relationship between data and graphic via continuous maps
  \item property preservation from data component to visual representation as monoid action equivariant maps
  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
\end{enumerate}

\section{Related Work}
%
% My best paraphrasing of the main idea would be that this paper takes the ideas of going from data to graphic along with some important constraints that people generally agree are important and formalizes them into a topological model, then shows how that model can map into a concrete implementation. The important part of this model is that it models the data sufficiently abstractly that it's able to work with basically any form or type of data. The sentence starting at line 489 seems to be the big "aha" to me, basically saying something like "if you implement things such that the code equivalents of these terms are separated in some specific way, then you get these important formally-describable properties". Is that right? If so, I'm left very curious about what alternative (possibly already-existing?) Artist implementations there might be that seem to work, but don't decompose things in this way, and therefore don't have these important properties. Or maybe the existing implementations do have these properties but aren't general enough? I know there are space constraints etc but if there was some way of showing how the current Matplotlib artists work and how they fall short of this model, that would be very enlightening I think.

@@

\subsection{Artist}
The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
\begin{equation}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
    \label{eq:math:artist:artist}
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
\begin{equation}
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation}
such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.

@@

\subsection{Limitations}
This model and the associated prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3. Even though the model is designed to be backend and format independent, it has only really been tested against PNGs rendered with the AGG backend. It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}. While this model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation. Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope for this model.

\subsection{Future Work}
More work is needed to formalize the composition operators and equivalence class \vartisteq.  More artists need to be implemented to demonstrate that the model can underpin a minimally viable library, foremost an image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as the boxplot\cite{wickham40YearsBoxplots2011}. Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices. While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

\section{Conclusion}
The data model and functional transform refactor presented here allows bulding block libraries to better support domain specific libraries without having to explicitly take in the specific data structure and visualization needs of those domains. Adopting this model would induce a separation of data representation and visual representation that for example in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base. Because of the model presented here, this refactor can be done in consistent and robust manner such that the resulting components can be be guaranteed to preserve continuity and structure as defined by the domain specific library developers without having to fold those definitions back into the base library.

%% if specified like this the section will be committed in review mode
\acknowledgments{
The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran. Also much appreciation to Nicolas Kruchten, Eric Schles, and Chana Tilevitz idenfiying the confusing bits and offering lots of suggestions. Thank you Marc Hanwell and Lev Manovich for helping refine the motivation, and Robert Haralick and Huy Vo for invaluable feedback on various iterations of this work. This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}

>>>>

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
%% this is still a mess, come back to it

Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built \cite{wongsuphasawatNavigatingWideWorld2021}. These building block libraries provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools; because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle. To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be a structure preserving map from data to visual representation.

\begin{figure}
  \includegraphics[width=\columnwidth]{figures/continuity.png}
  \caption{The line plot does not preserve continuity because it implies that the discrete records are connected to each other, while the bar plot is continuity preserving because it visually represents the records as independent data points.}
  \label{fig:intro:continuity}
\end{figure}

The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}. As illustrated in \autoref{fig:intro:continuity}, \textit{continuity} is the way in which the records in the dataset are connected to each other. While the visualization tool can generate the line plot, it does not preserve continuity structure because it implies that the discrete categories are somehow continuous. 
\begin{figure}
  \includegraphics[width=\columnwidth]{figures/equivariance.png}
  \caption{The data in blue is scaled by a factor of two, yielding the data in orange. To preserve equivariance, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
  \label{fig:intro:equivariance}
\end{figure}
Equivariance means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.

Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation. The contribution of this work is a  
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}

\section{Related Work}
The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner at a building building blocks library level. When introducing the retinal variables, Bertin informally introduces the notion that continuity is preserved in the mark and  defines equivariance constraints (selective, associative, ordered, quantitative)\cite{bertinSemiologyGraphicsDiagrams2011a}. In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another. The algebraic model of visualization proposed by Kindlmann and Scheidegger restrict equivarance to group actions, but explicitly define composable symmetric mappings from data space to representation space to graphic space. Our model targets monoid actions as they are more restrictive than all binary operations but more general than groups, and explicitly formalizes the way in which continuity is preserved as topologically maps. APT and the algebraic model are also targeted at visual designs and tools that automatically generate those designs, while our model targets the tools that build those tools.

Current visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}.  For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured. Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects\cite{loudenProgrammingLanguagesPrinciples2010}, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.

@@

\subsection{Artist}
\label{sec:math:artist}
The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
\begin{equation}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
    \label{eq:math:artist:artist}
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
\begin{equation}
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation}
such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.

@@

\subsection{Limitations}
Our model and prototype are deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3. Even though the model is designed to be backend and format independent, it has only been tested against PNGs rendered with AGG\cite{shemanarevAntiGrainGeometry}. It is unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection. While our model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation. Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope.

\subsection{Future Work}
More work is needed to formalize the composition operators and equivalence class \vartisteq and  we need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}. Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014}. While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

\section{Conclusion}
The data model and functional transform refactor presented here allows bulding block libraries to better support domain specific libraries without having to explicitly take in the specific data structure and visualization needs of those domains. Adopting this model would induce a separation of data representation and visual representation that, for example, in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base. A refactor driven by TEAM would result in components that can be be guaranteed to preserve continuity and structure, as defined by the domain specific library developers, without having to fold domain specific assumptions back into the base library.

%% if specified like this the section will be committed in review mode
\acknowledgments{
The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran, Eric Schles and Chana Tilevitz for idenfiying the confusing bits, Nicolas Kruchten for articulating the framing and Marc Hanwell, Lev Manovich, Robert Haralick and Huy Vo for invaluable feedback on various iterations. This paper is indebted to Kindlmann and Scheidegger's \cite{kindlmannAlgebraicProcessVisualization2014} for serving as a model and Munzner's guide to writing visualization papers. This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}
